---
title: "ESM 244 Lab"
author: "Chelsie McWhorter"
date: "1/6/2022"
output: 
    html_document:
    toc: yes
    toc_depth: 5
---

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(tidyverse)
library(here)
library(sf)
library(tmap)
#update.packages(ask = FALSE)
```
# Lab 1: Wrangling Refresher and Initial Map Building

### Read in the data

```{r}
sf_trees <- read_csv(here("data", "sf_trees","sf_trees.csv"),
                     show_col_types = FALSE)
```

### Example 1: wrangling and ggplot review

```{r}
## method 1: groupby() %>%  summarize()
method1 <- sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

##method 2: new functions
top5status <- sf_trees %>% 
  count(legal_status) %>%  #pseudo groupby function 
  drop_na(legal_status) %>%  #drops any row with an NA in it, so we want to specify to only drop legalsts
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% #default is to move it to the front
  slice_max(tree_count, n = 5) %>%  #pick out the highest up to what numberhere we'll pick 5
  arrange(-tree_count) #add - for highest to lowest

#these ^ get you to the same place
```


### Make a graph with the top 5 I just created:

```{r}
top5_graph <- ggplot(data = top5status, aes(x = fct_reorder(legal_status, -tree_count), y = tree_count)) + 
  geom_col(fill = 'darkgreen') +
  labs(x = "Legal Status", 
       y= "Tree Count") +
  coord_flip() + #flips the labels
  theme_minimal()

top5_graph #out of order, lets make them go biggest to smallest, not alphaberical order
```

### More Wrangling - Example 2: Keep only observatopns where legal stats is "Permitted Site" and caretaker is "MTA" 


```{r}

# sf_trees$legal_status %>% unique() #what options for legal status are there? are there any duplicates?

#unique(sf_trees$caretaker) #same as above but using a different command 

permitted_data <- sf_trees %>% 
  filter(legal_status == "Permitted Site", caretaker == "MTA") #the comma (or &) means both conditions have to be met in order to keep the line, in other words they BOTH have to be true. If we wanted one or the other than we used | for or
```

### Example 3: Only keep blackwood acacia trees, and then only keep columnc of legal status "date", "latitude", longitute, and store as blackwood_acacia

```{r}
blackwood_acaia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% 
  select(legal_status, date, lat = latitude, long = longitude)

###Make a graph of locations
ggplot(data = blackwood_acaia, aes(x = long, y = lat)) +
  geom_point(color = 'darkgreen') #currently R doesn't know its spatial, its just graphhing 
```

### Eample 4: Use `tidyr::separate` to break apart a column. Here we'll use the species column. 

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into =c('species_scientific', 'species_common'), sep = ' :: ')
```




### Example 5: `tidyr::unite` to bring things together


```{r}
ex5 <- sf_trees %>% 
  unite('id_status', tree_id, legal_status, sep = '_COOL_')

```

## Part 2: Make Some Maps

### Step 1: Covert lat & long to spatial points, using `st_as_sf()`

```{r}
##creating spatial object
blackwood_sf <- blackwood_acaia %>% 
  drop_na(lat, long) %>% 
  st_as_sf(coords = c('long', 'lat'))

### we need to tell r what hte coord red system is
st_crs(blackwood_sf) <- 4326 #epsg for WGS84

ggplot(data = blackwood_sf) +
  geom_sf(color = 'darkgreen') +
  theme_minimal()
```

Now R knows these are spatial data, and as such, we can now insert a map under these data to give it more context. 

### Insert a map into this spatial object. 

```{r}
sf_map <- read_sf(here('data', 'sf_map', "tl_2017_06075_roads.shp"))

# st_crs(sf_map) - this lets us look at the spatial data associated with an object 

sf_map_transform <- st_transform(sf_map, 4326)

#view it:
ggplot(data= sf_map_transform) +
  geom_sf()

```


### Combine map and data into one:

```{r}
ggplot() +
  geom_sf(data = sf_map,
          size = 0.1, 
          color = "darkgrey") +
  geom_sf(data = blackwood_sf, 
          color = "red", 
          size = 0.5) +
  theme_void() +
  labs(title = "Blackwood acacias in San Francisco")
```


### Now an interactive map:

```{r}
tmap_mode('view') #interactive

tm_shape(blackwood_sf) +
  tm_dots()
```



